Binning Strategies and Related Loss for Large Data

### 2015-02-27 Paper.Rnw
comments in section 2

All of these methods fall short in the example.
As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help some, but are as prone to over-plotting as solid circles when the data set is large. Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

@@

\ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points. A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into many segments on different scales. Theses approachs make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}

One approach to reduce the graphical complexity is to bin the displayed data. This has the additional advantage of reducing the size of the stored data, as only the bin centers and the frequency of points which correspond to that bin must be stored. \ktm{ Add \citep{Wickham2013Bin}  \citep{Liu2013imMens} here to argue the merits of binning. }

Histograms are an example of plots which use binned variables. To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.

@@

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion. These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.

As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e). The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c). The binned scatterplot is less prone to the Hermann-grid illusion %\hh{need a reference} 
than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.

>>>>

All of these methods fall short in the example.
As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help \hh{to a degree, but are also prone to over-plotting  when the data set is very large.} Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

@@

\ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales. \hh{These approaches do not help in the situation of ties in the data, which are very common for large data sets due to the finite resolution of data collection (e.g.\ airline data is collected at minute intervals).}  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points. A\hh{n} extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into many segments on different scales. Theses approachs make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.} \hh{please check the spelling in the previous paragraph. }

One approach to reduce the graphical complexity is to bin the displayed data. This has the additional advantage of reducing the size of the stored data, as only the bin centers and the frequency of points which correspond to that bin must be stored. \ktm{ Add \citep{Wickham2013Bin}  \citep{Liu2013imMens} here to argue the merits of binning. } \hh{also add \ref{gold} for binning reference - it's called `weighting' there.}

Histograms are an example of plots which use binned variables. \hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information. Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
%To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step
\hh{A natural extensions of histograms to higher dimensions is} to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, \hh{similar to} a two-dimensional histogram viewed from above.

@@

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion. \hh{However,} these estimates require \hh{careful} parameter tuning, \hh{as over-smoothing} may hide gaps in the data while simultaneously de-emphasizes outlying points.

As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) \hh{XXX the bubble plots already got mention above XXX} and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e). The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c). The binned scatterplot is less prone to the Hermann-grid illusion \hh{needs a reference} 
than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
